# Hypothesis testing {#sec-inference}

Hypothesis testing is a systematic way of answering questions about a population using data from a random sample. It is the backbone of statistical inference and essential to have an understanding for to be able to use more powerful techniques, like linear and logistic regression. In this chapter we will be working with data presented by the Swedish national public television broadcaster, SVT, on the current election poll (19/9-2023) [Väljarbarometern](https://www.svt.se/special/valjarbarometern/).

For a more technical/mathematical introduction to hypothesis testing, please consult the chapter on multiple testing (Chapter 13, 2nd edition) in [An Introduction to Statistical Learning](https://www.statlearning.com/).

<!-- [^inference]: Statistical inference can sometimes be tricky to understand. Essentially, statistical inference takes place whenever we learn something from data. We *infer* knowledge *from data*. -->

## Sample and population
We will be using the *sample* in Väljarbarometern (visited 19 September 2023) collected by Kantar Public by interviewing 3072 people between August 21 and September 3, 2023. We call this a sample since it is a small part of the whole *population* of all citizens of Sweden.

<span style="color: gray;">The data comes in aggregated form, i.e. in percentages, of the 3072 people but we have recreated the original sample below. Each row is the answer of one participant in the poll and the column `party` shows what the person said they would vote for. Below we show the code for recreating the sample from aggregated data. You don't need to understand the code, but if you want to recreate the sample yourself you can use the function `recreate_valjarbarometern`.</span>

```{r}
#| warning: false
#| code-fold: true
#| code-summary: "Code to recreate sample"
library(tidyverse)

recreate_valjarbarometern <- function(valjarbarometern_aggregate, n)
{
  # Calculate number of observations of each party
  obs_per_party <- round(n * valjarbarometern_aggregate$percentage / 100)
  
  # Create a list of party observations
  party_observations <- mapply(rep, valjarbarometern_aggregate$party, obs_per_party)
  
  # Flatten the list into a single vector
  sample_data <- unlist(party_observations)
  
  # Create a data frame with the samples
  valjarbarometern <- tibble(party = sample_data)
  
  return(valjarbarometern)
}
# Number of people in survey
n = 3072

parties <- c("V", "S", "MP", "C", "L", "KD", "M", "SD", "Other")
# Aggregated sample
valjarbarometern_aggregate <- tibble(
  party = factor(parties, levels = parties), # Using levels to preserve order
  percentage = c(7.6, 37.6, 4.3, 3.9, 3, 3.1, 19, 19.7, 1.7))

valjarbarometern <- recreate_valjarbarometern(
  valjarbarometern_aggregate, n)
```

We can take a look at the summary of our sample 

```{r}
summary(valjarbarometern)
```

The summary tells us how many people would vote for each party if the election were today, in our sample. A more common way to look at this is using a bar chart and showing the y-axis in percentage. The dashed line indicates the [limit](https://www.informationsverige.se/en/om-sverige/att-paverka-i-sverige/det-svenska-valsystemet.html) for being included in the Swedish parliament (Sveriges riksdag), 4%.

```{r}
# Define the colors in the order V, S, MP, C, L, KD, M, SD, Other. Make sure party column follows the same order!
party_colors <- c("#800000", "#FF0000", "#90EE90", "#008000", "#0054A6", "#00008B", "#00AEEF", "#FFFF00", "#808080")
valjarbarometern |>
  ggplot(aes(x = party, fill = party)) + 
  geom_bar(aes(y = after_stat(count/sum(count) * 100))) + # Calculating percentages
  geom_hline(
    yintercept = 4,
    col = "black",
    linetype = 2) +
  labs(
    x = "Party",
    y = "Percentage") +
  guides(fill = "none") + # Remove fill legend
  scale_fill_manual(values = party_colors) # Set colors to party colors
```

In our bar chart we see that several parties seems to be at risk of ending up below 4% if there was an election today: MP (Miljöpartiet) seems to be just above the 4% limit and C (Centern) just below, while L (Liberalerna) and KD (Kristdemokraterna) seem worst off. Let's take a closer look at the percentages

```{r}
n_participants <- 3072 # Number of participants in survey
valjarbarometern |>
  summarize(
    percentage = round(100 * n() / n, digits = 1), # Calculating percentage
    .by = party) # Grouping by party
```
  
The summary confirms what we could tell from the bar chart, MP is just above the 4% limit with 4.3% and C is just below the limit with 3.9%. But this result comes from a sample of 3072, what does that actually tell us about how the whole *population* of Sweden would vote? We would like to know what the value would be if the whole population voted. To learn about this from our sample, we need to perform statistical inference, and in particular hypothesis testing.


### Exercises

1. Previously we worked with the `palmerpenguins` dataset. What is the population in that dataset? Use `?penguins` after loading the package (`library(palmerpenguins)`) to read more about the data. Where and when was the data collected? 

## Hypothesis testing (proportions)
Hypothesis testing is method that allows us to draw inferences about a population using a sample. In more simpler terms, hypothesis testing let's us use a sample to answer questions about the population with some uncertainty. The procedures follows four steps: 

**1. Formulate hypotheses**

Formulate the hypotheses as a null hypothesis, $H_0$, and an alternative hypothesis, $H_a$ (sometimes $H_1$). The hypotheses should reflect the question that we have: in our case, for MP, we want to know if the percentage in the population is larger than 4%. We can formulate the hypotheses for this question as

> **$H_0$**: The percentage of people who would vote for MP is *smaller than or equal* to 4%.

> **$H_a$**: The percentage of people who would vote for MP is *larger* than 4%.

$H_0$ and $H_a$ play different roles in hypothesis testing, $H_0$ is our baseline and $H_a$ represents the hypothesis we want to prove. The hypotheses should be mutually exclusive and exhaustive, that is they should cover every possible outcome. In our example the percentage must be a number between 0 and 100, which must be either smaller than or equal to 4 or larger than 4, i.e. our hypotheses cover all possibilities.

<span style="color: gray;">**2. Calculate test statistic**</span>

<span style="color: gray;">Calculate a test statistic that summarizes the *evidence* in our data *against* the null hypothesis. This part is were most of mathematical and statistical theory is needed and **in this course we will leave this to a test function in R.** Instead we will run use a test funciton in R.</span>

**2. Use a test function in R**

Run a function in R that performs a statistical hypothesis test that summarizes the *evidence* in our data *against* the null hypothesis.

**3. $p$-value**

Calculate a $p$-value based on the evidence, which is done by the test function. The $p$-value is of great importance in hypothesis testing and it is important to have a basic understanding of it. In our case, the $p$-value is the probability of observing 4.3% or larger (further from the null hypothesis) in a sample of 3072 people, *if the null hypothesis was true*. $p$-values can always be formulated in terms of the null hypothesis and the observed parameter, in this case the percentage voting for MP, as: "The probability of observing the *observed value* or more extreme, i.e. 4.3% or larger, in a sample given that the null hypothesis is true".  

**4. Decision**

Decide to reject or not reject the null hypothesis based on the $p$-value. This is a decision rule that is decided beforehand, based on a *significance level*, usually denoted $\alpha$. If the $p$-value is smaller than the significance level, we reject the null hypothesis at significance level $\alpha$. The most common significance level is 0.05, and we will use it throughout this course. If the $p$-value is *smaller* than 0.05 we say that the evidence in the data is *statistically significant* at *significance level*, $\alpha$, 0.05 to reject the null hypothesis. This decision rule is dichotomous, i.e. it is either true or false. This is commonly used, but the $p$-value is valuable on it's own as a $p$-value of 0.06 is still a very interesting result, while a $p$-value of 0.9 is not. A low $p$-value may indicate that "there may be something there", even if it is not significant at significance level 0.05.

In practice we will not spend too much time on hypothesis testing alone in this course, but instead focus more on regression analysis where hypothesis testing is an important component. In fact, as you will see, regression analysis can be used for hypothesis testing.

### Testing for significance
So how about MP and C? Is the percentage of people saying they would vote for MP over 4% with statistical significance? Let's perform a hypothesis test. Note that we will be using *proportions* and not percentages, but it is a simple conversion, proportion = percentage / 100. Formulated in proportions our hypotheses are

> **$H_0$**: The proportion of people who would vote for MP is *smaller* than 0.04.

> **$H_a$**: The proportion of people who would vote for MP is *larger* than 0.04.

```{r}
n_participants <- 3072 # Number of participants in survey
n_mp_voters <- valjarbarometern |> # Number of participants who answered MP
  filter(party == "MP") |>
  summarize(n = n()) |>
  pull(n)
p_h0 <- 0.04 # Our null hypothesis
# Conducting hypothesis test of proportions. Observe that we are testing if the proportion is *greater* than.
result <- prop.test(n_mp_voters, n_participants, p_h0, alternative = "greater")
result
```

The result of our test contains a lot of information, but we only need to focus on the part that says "p-value" on the second line and read its value, rounded 0.21. Note that we should *not* read the p at the second to last line, as it is the proportion we observed in our sample. We compare the $p$-value with our significance level, 0.05 and conclude that 0.21 is *larger* than 0.05, and thus we *can not* reject our null hypothesis. This means we *do not* have evidence in the data that the percentage (proportion) of Sweden's population that would vote for MP is larger than 4% (0.04), i.e. we cannot reject the null hypothesis. So the conclusion of our test is that if there was an election today MP would not be "statistically significantly" safe. 

We can also access the $p$-value directly by accessing the `p.value` using the dollar operator, `$`, from our result.

```{r}
result$p.value
```

### Exercises
1. How would the hypotheses change if we were testing if the percentage that would vote on the C party is less than 4%?
2. Perform a hypothesis test to see if the percentage that would vote on C is statistically significant *less* than 4%. Reuse the code from the chapter!
3. Find a party that is below the 4% limit with statistical significance at significance level 0.05. Try yourself or expand the code below to see the answer. Reuse the code from the chapter!

## Comments on hypothesis testing
A few important things to note in general about hypothesis testing:

* We look for evidence *against* the null hypothesis $H_0$, but failing to reject the null hypothesis doesn't prove the null hypothesis is true. It only suggests that there isn't enough statistical evidence against it.
* A significantly small p-value doesn't mean the difference between the mean under $H_0$ and the actual population mean is large, only that it's likely not due to chance. 
* The significance level should be set *beforehand*, not changed when the $p$-value is observed. Usually 0.05 is used in scientific analysis.
* The significance level 0.05 is a strong indicator to reject the null hypothesis, but sometimes it may be too black and white. A $p$-value of 0.06 or 0.1 still indicates an interesting result, while a $p$-value of 0.9 does not. The $p$-value offers the opportunity for the reader to decide by themselves if they find the result interesting and worth further studies.
* In practical applications, the outcome of hypothesis testing should be combined with domain knowledge and other information to make informed decisions or answer research questions. 

<span style="color: gray;">
When performing hypothesis testing two different errors can be committed and if you want to further study hypothesis testing and/or statistics it is important to know them.

* Type I Error: When you incorrectly reject a true null hypothesis. The probability of committing this error is $\alpha$, the significance level.
* Type II Error: When you fail to reject a false null hypothesis. The probability of committing this error is denoted by $\beta$. The power of a test, $1 - \beta$, is the probability of rejecting a false null hypothesis.
</span>

## Hypothesis test (mean)
Up until now we have studied hypothesis test of proportions. A perhaps more common test is hypothesis test of the mean and difference in means. For this, we can look the `penguins` dataset again and specifically the variable `body_mass_g`, the weight of the penguins. Before we start looking at the variable we should think about our sample and population. The data was collected during the years 2007-2009 in the Palmer Archipelago, Antarctica. If we believe that the dataset is still valid for penguins in the Palmer archipelago today, we can say that our population is all penguins of the species Adelie, Chinstrap, and Gentoo in the Palmer archipelago. Now when we have identified the population we are studying, we can start by looking at the variable. We will start by visualizing the distribution over body mass and the mean body mass marked by a dashed line.

```{r}
#| warning: false
library(palmerpenguins)
penguins |> 
  ggplot(aes(x = body_mass_g)) + 
  geom_density() + 
  geom_vline(
    xintercept = mean(penguins$body_mass_g, na.rm = TRUE),
    linetype = 2) +
  labs(
    x = "Body mass (g)",
    y = "") + 
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank())
```

We see that there are quite a lot of observations with a smaller body mass than the mean. Can we be sure that the population mean, i.e. the mean of all penguins body weights, including those outside our sample, is above 4000 g? To test this we use a t-test. The t-test can be used we we want to ask questions about the population mean based on our sample. We formulate our hypothesis

> **$H_0$**: The average body mass of all penguins in the Palmer Archipelago is 4000 g.
 
> **$H_a$**: The average body mass of all penguins in the Palmer Archipelago is *more* than 4000 g.

We can perform this test in R using the `t.test` function.

```{r}
t.test(penguins$body_mass_g, mu = 4000, alternative = "greater")
```

Again we are only interested in the $p$-value, and if it is smaller than 0.05. We see that "p-value = 2.35e-06". Numbers with an "e" is represented using *scientific notation*, and it means $2.35 \cdot 10^{-06} = 0.000000235$, which is a very, *very* small number. In general the number after e can be understood as the number of zeros before the first significant digit, e.g. 1.2e-02 = 0.0012. So we can conclude that at significance level 0.05 we have evidence that the mean penguin weight is more than 4000 since our $p$-value is smaller than 0.05.

## Difference in means
Using your skills from EDA (or memory of the Data visualization chapter in R4DS), you may have noticed that the density plot looked a bit strange: it looks like there are several densities in our sample, noticeable by the humps at around 3600, 4700, and 5500 grams. Let's explore this further, and a good place to start is one of the categorical variables that may indicate different groups. We can visualize the density again, this time coloring the fill after the different species.

```{r}
# warning: false
penguins |>
  drop_na() |>
  ggplot(aes(x = body_mass_g, fill = species)) +
  geom_density(alpha = 0.3) +
  labs(
    x = "Body mass (g)",
    y = "",
    fill = "Species") + 
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank())
```

That was revealing! But still there seems to be something going on... If we look at the distribution over Gentoo penguins, we see two distinct humps again. Another categorical variable that we may suspect will affect the body mass is sex. We take a closer look at only the Gentoo penguins and color the fill based on sex instead.

```{r}
library(palmerpenguins)
penguins |>
  drop_na() |>
  filter(species == "Gentoo") |>
  ggplot(aes(x = body_mass_g, fill = sex)) +
  geom_density(alpha = 0.3) +
  labs(
    x = "Body mass (g)",
    y = "",
    fill = "Sex") + 
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank())
```

We see a clear difference, but will this difference hold in the population of Gentoo penguins? How big is our sample of Gentoo penguins?

```{r}
penguins |>
  filter(species == "Gentoo") |>
  drop_na() |>
  summarize(
    count = n(),
    .by = sex)
```

We have 58 female Gentoos and 61 males, it is not that few, but according to [Wikipedia](https://en.wikipedia.org/wiki/Gentoo_penguin) the population of Gentoos is more than 600,000 birds! Is this difference statistically significant? Again we can use a t-test to compare the mean of two groups. We formulate our hypothesis

> **$H_0$**: Female and male Gentoos have the same body mass on average.

> **$H_a$**: Female and male Gentoos *do not* have the same body mass on average.

We use `t.test` again to test for statistical significance.

```{r}
female_gentoo_body_mass <- penguins |>
  filter(species == "Gentoo" & sex == "female") |>
  select(body_mass_g) |>
  drop_na()
male_gentoo_body_mass <- penguins |>
  filter(species == "Gentoo" & sex == "male") |>
  select(body_mass_g) |>
  drop_na()

t.test(male_gentoo_body_mass, female_gentoo_body_mass)
```

We see that the $p$-value is very small, $2.2 \cdot 10^{-16}$. We conclude that we have evidence that the difference in average body weight of female and male Gentoo penguins is statistically significant at significance level 0.05. 

Now we have looked at two different ways of performing hypothesis testing about different variables, either proportions or means. In the next chapter we will look at *linear regression*, a method that we can use to understand if there is an effect of one variable on another and to make powerful predictions.

### Exercises

1. Test if the mean flipper length is more than 180 mm. Try to write down the hypotheses before you perform your test. Is the result statistically significant?
2. Test if the mean flipper length differs between species. Formulate the hypothesis and draw conclusions from the tests. Based on your result, is flipper length a good way to distinguish between different penguins species? You can use the code below to get started, or try yourself first.
```{r}
#| code-fold: true
#| code-comment: "Hint"
#| execute: false
gentoo_flipper_length <- penguins |>
  filter(species == "Gentoo") |>
  select(flipper_length_mm) |>
  drop_na()
chinstrap_flipper_length <- penguins |>
  filter(species == "Chinstrap") |>
  select(flipper_length_mm) |>
  drop_na()
adelie_flipper_length <- penguins |>
  filter(species == "Adelie") |>
  select(flipper_length_mm) |>
  drop_na()
```
3. Try to figure out how to perform a proportion test to study if there are more Chinstrap penguins than Adelie on Dream island. Start by visualizing the distribution using some visualization of Island and Species. Then you will need to count the number of Chinstrap penguins on Dream island as well as the total number of penguins on Dream island. Do you see any problem with the sample we are using for this test? You can use the code below as a hint.
```{r}
#| code-fold: true
#| code-comment: "Hint"
#| execute: false
# Visualization
penguins |> 
  ggplot(aes(x = island, fill = species)) + 
  geom_bar()
# Total number of penguins on Dream island
n_total <- penguins |> 
  filter(island == "Dream") |>
  summarize(n = n()) |> 
  pull()
# Number of chinstrap penguins on Dream island
n_chinstrap <- penguins |> 
  filter(species == "Chinstrap" & island == "Dream") |>
  summarize(n = n()) |> 
  pull()
```
