# R4DS: Data tidying

To run these solutions you need to load the necessary libraries.

```{r}
library(tidyverse)
```

## 6.2 Tidy data
### Exercise 1

* **table1**
  - Observation: Tuberculosis and population for one country and one year.
  - Columns:
    - country, year, tuberculosis cases, population
* **table2**
  - Observation: Tuberculosis or population for one country and one year.
  - Columns:
    - country, year, type of observation, count of the type of observation
* **table3**
  - Observation: Tuberculosis per capita as a string for a country and one year.
  - Columns:
    - country, year, tuberculosis per capita as a string

Observe that the rate in table three is stored as a string, which is not the value of tuberculosis per capita.

### Exercise 2

Not recommended, but here is a solution for table2, if you are interested. You do not need to be able to reproduce this.

```{r}
#| code-fold: true
#| code-comment: Code
table2 <- tibble(
  country = c(
      "Afghanistan",
      "Afghanistan",
      "Afghanistan",
      "Afghanistan",
      "Brazil",
      "Brazil"
  ),
  year = c(1999, 1999, 2000, 2000, 1999, 1999),
  type = c(
      "cases",
      "population",
      "cases",
      "population",
      "cases",
      "population"
  ),
  count = c(745, 19987071, 2666, 20595360, 37737, 172006362)
)

cases <- table2 |> 
  filter(type == "cases") |>
  pull(count)
population <- table2 |> 
  filter(type == "population") |>
  pull(count)
rate <- (cases / population) * 10000
table2$rate <- rep(rate, each = 2)
table2
```
