# Introduction to R

## Objects

### Exercise 1
In R it is possible to use `=` and `<-` to assign values to objects, but `<-` is the preferred way. `=` is used to assign function parameters a value, which you will encounter later.

## Functions

### Exercise 1
```{r}
multiply <- function(x, y) {
  x * y
}

multiply(36, 52)
```

### Exercise 2

Make sure you know which variable appears on what sign of the minus sign when using the function.

```{r}
subtraction <- function(x, y) {
  x - y
}

subtraction(36, 52)
```

### Exercise 3

Here we see the usage of `=` in assigning function parameters values. 

```{r}
subtraction(y = 52, x = 36)
```

### Exercise 4

```{r}
subtraction(y <- 52, x <- 36)
```

Instead of passing x and y as parameters to the function, `<-` sets the values of the objects `x` and `y` to 36 and 52 respectively, which we can see by printing the variables: 

```{r}
x
y
```

The output is the same as the following code

```{r}
x <- 36
y <- 52

subtraction(y, x)
```

**Never use `<-` inside a function call.**

## Vectors

### Exercise 1

The operations, $+, -, \cdot, /$, are performed $elementwise$, that is the first element of bot vectors gets added/subtracted/multiplied/divided and so on. 

```{r}
vec1 <- c(1, 2)
vec2 <- c(3, 4)

vec1 + vec2
vec1 - vec2
vec1 * vec2
vec1 / vec2
```

### Exercise 2

We try running the code:

```{r}
#| error: true 
vec1 <- c(1, 2)
vec2 <- c(3, 4)
vec3 <- c(3, "four")

vec1 + vec2
vec2 + vec3
```

We get an error for `vec2 + vec3`. Vectors must always contain the same data type, so in this case `vec3` becomes a vector of strings:

```{r}
vec3
```

We can see that 3 is now also a string by the quotes, `"3"`.

## Data frames

### Exercise 1

We run the code and see that the output is similar to the `head` function.

```{r}
# We comment the installation to not run it twice on our machine
# install.packages("palmerpenguins")
library(palmerpenguins)
penguins
```

### Exercise 2

We type the dollar operator straight after the tibble and then the column name to print the values in it.

```{r}
penguins$flipper_length_mm
```

### Exercise 3

When we press tab we see all the columns in the tibble as suggestions. In general you can use tab to autocomplete in R.

### Exercise 4

We see some output when loading `tidyverse` which you can ignore for now. If you have questions about the functions, run `?<function>`, e.g. `?glimpse` in the console.

```{r}
library(tidyverse)
glimpse(penguins)
summary(penguins)
head(penguins)
tail(penguins)
nrow(penguins)
ncol(penguins)
names(penguins)
```
