# Multiple logistic regression

## Setup 

First we load the dataset and have a `glimpse` at it.

```{r}
#| warning: false
library(tidyverse)
#install.packages("mlbench") # Install to get PimaIndiansDiabetes2 dataset
library(mlbench)
data("PimaIndiansDiabetes2") # Using 2 because it is a corrected version.
glimpse(PimaIndiansDiabetes2)
```

## Model selection
### Exercise 1

We fit the two models and use the `AIC` function

```{r}
# Pressure model
model <- glm(diabetes ~ pressure, data = PimaIndiansDiabetes2, family = "binomial")
AIC(model)
# Triceps model
model <- glm(diabetes ~ triceps, data = PimaIndiansDiabetes2, family = "binomial")
AIC(model)
```

We see that `triceps` has much lower AIC and comparing to the result in the lecture notes including `pressure` barely improves the model!

### Exercise 2

We fit the model with all three predictors and look at the AIC value and the VIF value

```{r}
model <- glm(diabetes ~ pressure + mass + triceps, data = PimaIndiansDiabetes2, family = "binomial")
summary(model)
```

We see that the AIC is lower than using only `pressure` and `triceps`, but the $p$-value for `triceps` is now over 0.05, indicating that the relationship is not statistically significant at significance level 0.05. However, the $p$-value is low so we can keep it in our model, but it may be reasonable to use only one of BMI and tricep skin fold thickness as both are measurements relating to body fat.

We also calculate the VIF value to see if we have multicollinearity

```{r}
# warning: false
library(car)
vif(model)
```

We see that all values are below 5, so we do not need to remove any variable because of multicollinearity.

### Model selection strategies
#### Exercise 1
See Assignment 2.

## Accuracy

### Exercise 1

We see that the accuracy is the same, with or without `pressure`. In this case the simpler model, not including `pressure` is the better option.

```{r}
model <- PimaIndiansDiabetes2 |> 
  glm(formula = diabetes ~ ., family = "binomial")

# Calculating the accuracy
PimaIndiansDiabetes2 |> 
  # Calculated the predicted probabilities for all observations
  mutate(predicted_prob = model |> predict(PimaIndiansDiabetes2, type = "response")) |> 
  # Predicting the outcome
  mutate(predicted_outcome = ifelse(predicted_prob > 0.5, "pos", "neg")) |> 
  # Dropping predictions with missing values
  filter(!is.na(predicted_outcome)) |>
  # Summarizing (Accuracy = # Correct predictions / # Total predictions
  summarize(accuracy = mean(predicted_outcome == diabetes))
```

### Exercise 2

We leave this one to you.

### Exercise 3

Using this seed the accuracy is the same. 

```{r}
# code-fold: true
# comment: Code
# Set seed for reproducibility
set.seed(42) 

# Create an index to keep track of subjects
pima <- PimaIndiansDiabetes2 |> 
  drop_na() |> # Using only complete observations
  mutate(index = 1:n())

# Create training and testing set
training_set <- pima |> 
  sample_frac(0.8) # Use 80% of the data in training set
test_set <- pima |> 
  anti_join(training_set, by = "index")

# Train model on training set
model <- training_set |> 
  glm(formula = diabetes ~ ., family = "binomial")

# Evaluate model by accuracy on test set
predictions <- model |> 
  predict(test_set, type = "response")
test_set |> 
  mutate(prediction = ifelse(predictions > 0.5, "pos", "neg")) |> 
  drop_na() |> # Dropping predictions with missing values
  summarize(accuracy = sum(prediction == diabetes) / n())

model <- training_set |> 
  glm(formula = diabetes ~ . - pressure, family = "binomial")

# Evaluate model by accuracy on test set
predictions <- model |> 
  predict(test_set, type = "response")
test_set |> 
  mutate(prediction = ifelse(predictions > 0.5, "pos", "neg")) |> 
  drop_na() |> # Dropping predictions with missing values
  summarize(accuracy = mean(prediction == diabetes))
```

With a different seed the full model has a higher accuracy.

```{r}
# code-fold: true
# comment: Code
# Set seed for reproducibility
set.seed(44) 

# Create an index to keep track of subjects
pima <- PimaIndiansDiabetes2 |> 
  drop_na() |> # Using only complete observations
  mutate(index = 1:n())

# Create training and testing set
training_set <- pima |> 
  sample_frac(0.8) # Use 80% of the data in training set
test_set <- pima |> 
  anti_join(training_set, by = "index")

# Train model on training set
model <- training_set |> 
  glm(formula = diabetes ~ ., family = "binomial")

# Evaluate model by accuracy on test set
predictions <- model |> 
  predict(test_set, type = "response")
test_set |> 
  mutate(prediction = ifelse(predictions > 0.5, "pos", "neg")) |> 
  summarize(accuracy = sum(prediction == diabetes) / n())

model <- training_set |> 
  glm(formula = diabetes ~ . - pressure, family = "binomial")

# Evaluate model by accuracy on test set
predictions <- model |> 
  predict(test_set, type = "response")
test_set |> 
  mutate(prediction = ifelse(predictions > 0.5, "pos", "neg")) |> 
  summarize(accuracy = sum(prediction == diabetes) / n())
```

To properly study accuracy it is good to use more than one train-test combination or using k-fold validation [cross-validation]('https://en.wikipedia.org/wiki/Cross-validation_(statistics)'). However, this is beyond the scope of this course. 

We see that including `pressure` doesn't seem to improve our accuracy much, suggesting it is better to go for the smaller, less complex model.

### Exercise 4

The model would achieve an accuracy of 90% since $\text{Accuracy} = \#\text{Correct predictions} / \#\text{Total predictions} = 90 / 100 = 0.9 = 90 \%$.
