# Linear regression

## Setup 

First we load the dataset and have a `glimpse` at it.

```{r}
#| warning: false
library(tidyverse)
library(palmerpenguins)
glimpse(penguins)
```

## A linear relationship
### Exercise 1

We see that the average body weight for a penguin with flipper length of 201 mm is 4206, an increase of 50 g from 200. Note we round the values in the plot outputs.

```{r}
#| warning: false
x <- 201
model <- penguins |> 
  lm(formula = body_mass_g ~ flipper_length_mm)
prediction <- data.frame(
  flipper_length_mm = x,
  body_mass_g = model |> predict(data.frame(flipper_length_mm = x)))

penguins |> 
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) + 
  geom_segment(x = prediction$flipper_length_mm, xend = prediction$flipper_length_mm,  y = 0, yend = prediction$body_mass_g, color = "black", linetype = 2, linewidth = 0.2) +
  geom_segment(x = 0, xend = prediction$flipper_length_mm,  y = prediction$body_mass_g, yend = prediction$body_mass_g, color = "black", linetype = 2, linewidth = 0.2) +
  scale_x_continuous(breaks = c(prediction$flipper_length_mm)) +
  scale_y_continuous(breaks = round(c(prediction$body_mass_g))) +
  geom_point(data = prediction, col = "red", size = 3) + 
  labs(
    x = "Flipper length (mm)", 
    y = "Body mass (g)")
```

### Exercise 2

The predicted average body mass is -812 g, which doesn't seem to make any sense! When we are predicting based on values that are outside of our sample we may get poor predictions, as in this case. **Always be mindful when predicting outside your sample.**

```{r}
#| warning: false
x <- 100
model <- penguins |> 
  lm(formula = body_mass_g ~ flipper_length_mm)
prediction <- data.frame(
  flipper_length_mm = x,
  body_mass_g = model |> predict(data.frame(flipper_length_mm = x)))

penguins |> 
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) + 
  geom_segment(x = prediction$flipper_length_mm, xend = prediction$flipper_length_mm,  y = 0, yend = prediction$body_mass_g, color = "black", linetype = 2, linewidth = 0.2) +
  geom_segment(x = 0, xend = prediction$flipper_length_mm,  y = prediction$body_mass_g, yend = prediction$body_mass_g, color = "black", linetype = 2, linewidth = 0.2) +
  scale_x_continuous(breaks = c(prediction$flipper_length_mm)) +
  scale_y_continuous(breaks = round(c(prediction$body_mass_g))) +
  geom_point(data = prediction, col = "red", size = 3) + 
  labs(
    x = "Flipper length (mm)", 
    y = "Body mass (g)")
```

## Fitting a linear model
### Exercise 1

```{r}
model <- penguins |> 
  lm(formula = flipper_length_mm ~ body_mass_g)
summary(model)
```

The linear model doesn't care which variable is the dependent and independent variable, it will fit a model either way. The coefficients will be very different, as the increase in body mass is much smaller per mm flipper length than the other way around, but the relationship will still be significant. The coefficient is 1.528e-02 which is the same as $1.528 \cdot 10^{-2} = 0.01528$. Thus the flipper length increases on average by 0.015 cm per g the body mass increases. We can never draw any causal conclusions without additional assumptions, and in this case it doesn't seem appropriate, as it seems more like a "chicken and egg" dilemma than a causal effect. The model can still be useful for predicting and to understand the relationship between the variables, without suggesting a causal relationship.

```{r}
#| warning: false
penguins |> 
  ggplot(aes(x = body_mass_g, y = flipper_length_mm)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) + 
  labs(
    x = "Flipper length (mm)", 
    y = "Body mass (g)")
```

### Exercise 2

From the visualization we see that the sales increase during the summer months and decreases during winter, but the line is completely flat. There is a clear relationship, but it is not a linear relationship! We can see from how the data is generated that it depends on the `sin` function, which is not linear.

```{r}
# Simulating data
day <- seq(0, 12, length.out = 356)
sales <- 400 * sin(pi * day / 12)
ice_cream <- tibble(day = day, sales = sales)

ice_cream |> 
  ggplot(aes(x = day, y = sales)) + 
  geom_point() + 
  geom_smooth(method = "lm") + 
  scale_x_continuous(
    breaks = seq(0, 12)) +
  labs(
    x = "Month",
    y = "Ice cream sales")
```

We see that the $p$-value is 1, there is no significant relationship. It is important to remember that we are looking for a *linear* relationship when performing linear regression. It will not detect any other relationship.

```{r}
ice_cream |> 
  lm(formula = sales ~ day) |> 
  summary()
```

### Exercise 3

We see that we get one coefficient for Chinstrap and one for Gentoo. The value, under "Estimate" is the difference from the Adelie penguins in average body weight, compare with the values computed below.

```{r}
penguins |> 
  lm(formula = body_mass_g ~ species) |>
  summary()
```

We can check this

```{r}
# Calculating mean body weight for each species
adelie_mean <- penguins |> 
  filter(species == "Adelie") |> 
  pull(body_mass_g) |>
  mean(na.rm = TRUE)
chinstrap_mean <- penguins |> 
  filter(species == "Chinstrap") |> 
  pull(body_mass_g) |>
  mean(na.rm = TRUE)
gentoo_mean <- penguins |> 
  filter(species == "Gentoo") |> 
  pull(body_mass_g) |>
  mean(na.rm = TRUE)

 # Differences in mean
chinstrap_mean - adelie_mean
gentoo_mean - adelie_mean
```

In addition to comparing the means the linear regression outputs $p$-values that tests if there is a significant difference between the species. Only the difference between Adelie and Gentoos is significant. 

### Exercise 4

We see that the $p$-values are the same and we can use linear regression instead of the t-test. Linear regression is more versatile and can accommodate many levels in the categorical predictor. If it is possible to use a t-test it may be more straightforward and clear though.

```{r}
# t-test
female_gentoo_body_mass <- penguins |>
  filter(species == "Gentoo" & sex == "female") |>
  select(body_mass_g) |>
  drop_na()
male_gentoo_body_mass <- penguins |>
  filter(species == "Gentoo" & sex == "male") |>
  select(body_mass_g) |>
  drop_na()

t.test(male_gentoo_body_mass, female_gentoo_body_mass)

# Linear regression
penguins |> 
  filter(species == "Gentoo") |>
  lm(formula = body_mass_g ~ sex) |>
  summary()
```


## Categorical predictors
### Exercise 1

We start with the linear model.

```{r}
penguins |> 
  lm(formula = body_mass_g ~ island) |> 
  summary()
```

We see that the coefficients for both Dream and Torgersen are significant, but let's have a look at a density plot to better understand the result.

```{r}
#| warning: false
#| code-fold: true
#| comment: "Code"
mean_bws <- penguins |> 
  summarize(
    body_mass_g = mean(body_mass_g, na.rm = TRUE),
    .by = island)
penguins |> 
  ggplot(aes(x = body_mass_g, fill = island)) + 
  geom_density(alpha = 0.3) +
  geom_vline(
    aes(
      xintercept = body_mass_g,
      col = island),
    data = mean_bws) +
  labs(
    x = "Body mass (g)",
    y = "",
    col = "Species",
    fill = "Species") +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank())
```

We see that `lm` choose Biscoe as the reference island, making the coefficients for both the other islands significant. The result would have been different if Dream or Torgersen would've been the reference. 

### Exercise 2

We set Torgersen as reference and rerun the linear regression.

```{r}
#| eval: false
penguins_re <- penguins |> mutate(island = relevel(island, ref = "Torgersen"))
model_summary <- penguins_re |> 
  lm(formula = body_mass_g ~ island) |> 
  summary()
model_summary$coefficients
```

We see that now only Biscoe is significant.

### Exercise 3

Gentoos only exists on Biscoe, which is the island with the significant difference in body weight. It is fair to assume that species is more responsible than island for the effect we see on body mass.

```{r}
penguins |> 
  ggplot(aes(x = island, fill = species)) +
  geom_bar() +
  labs(
    x = "Island",
    y = "Count",
    fill = "Species")
```

### Exercise 4

We see that island doesn't have a significant effect on the Adelie penguins body mass and the coefficients are quite small in comparison to the total body mass of the average penguin. 

```{r}
model_summary <- penguins |> 
  filter(species == "Adelie") |>  
  lm(formula = body_mass_g ~ island) |> 
  summary()
model_summary$coefficients
```

We can visualize the result to see that it seems to make sense.

```{r}
mean_bws <- penguins |> 
  filter(species == "Adelie") |>  
  summarize(
    body_mass_g = mean(body_mass_g, na.rm = TRUE),
    .by = island)
penguins |> 
  filter(species == "Adelie") |>  
  ggplot(aes(x = body_mass_g, fill = island, col = island)) + 
  geom_density(alpha = 0.3) +
  geom_vline(
    aes(xintercept = body_mass_g,
        col = island),
    data = mean_bws) +
  labs(
    x = "Body mass (g)",
    y = "",
    col = "Island",
    fill = "Island") +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank())
```

## A surprising result (Simpson's paradox)
### Exercise 2 

We see that the $p$-values are low for both species, even if it is just above 0.05 for Gentoos. Even if the points seemingly line up well, the difference is still statistically significant, at least for one species. This shows that it is hard to determine statistical significance by eye only.

```{r}
model_summary <- penguins |> 
  lm(formula = body_mass_g ~ flipper_length_mm + species) |> 
  summary()
model_summary$coefficients
```

```{r}
penguins |> 
  ggplot(aes(x = flipper_length_mm, y = body_mass_g, col = species)) +
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Flipper length (mm)",
    y = "Body mass (g)",
    col = "Species")
```



## Prediction and prediction interval
### Exercise 1

```{r}
#| warning: false
model <- penguins |>  lm(formula = body_mass_g ~ flipper_length_mm)
model |> predict(data.frame(flipper_length_mm = 201), interval = "prediction", level = 0.95)
model |> predict(data.frame(flipper_length_mm = 100), interval = "prediction", level = 0.95)
```

### Exercise 2

The interval becomes smaller. The confidence level specifies our confidence in a new observation falling within the interval. If we make the interval smaller, we have to reduce our confidence. On the other hand, if we make our interval infidelity big, we can be 100% confident it will cover the next observation. Of course there is no use for such a 100% interval in practice.

```{r}
#| warning: false
model |> predict(data.frame(flipper_length_mm = 200), interval = "prediction", level = 0.5)
```
